One-arm spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-06 paper/model.tex
minor edits

\subsection{Three-dimensionality}
In 3D simulations we measure the importance of vertical motion with
$\Theta$, where 
\begin{align}
  \Theta^2 \equiv \frac{\avg{v_z^2}}{\avg{v_R^2}+\avg{v_\phi^2}}, 
\end{align}
and $\avg{v_z^2}$ denotes the density-weighted average,
\begin{align}
  \avg{v_z^2} \equiv\frac{
    \int_{R_1}^{R_2}\int_{\theta_\mathrm{min}}^{\pi/2} \int_{0}^{2\pi}
    \rho v_z^2 dV}{
    \int_{R_1}^{R_2}\int_{\theta_\mathrm{min}}^{\pi/2} \int_{0}^{2\pi}
    \rho dV
  },
\end{align}
and similarly for the horizontal velocities. The radial range of
integration is taken over $r\in[R_1,R_2]$ since this is where
we find the perturbations to be confined.

@@

In a linear analysis, one assumes a steady axisymmetric background state, such as
that defined in \S\ref{model}, which is then perturbed such that
\begin{align}  
  \Sigma \to \Sigma(R) + \delta\Sigma_m(R)\exp{\left[\ii\left(-\sigma t +
        m\phi\right)\right]}, 
\end{align}
and similarly for other variables. Since the background is
axisymmetric, we identify $\delta\Sigma_m(R)\exp{\left(-\ii\sigma
    t\right)} = \Sigma_m$ as defined for diagnostics. 
The linearised mass and momentum equations are
\begin{align}
  &-\ii\sbar \delta\Sigma_m = -\frac{1}{R}\frac{d}{dR}\left(R\Sigma\delta
    v_{Rm}\right) - \frac{\ii m \Sigma}{R}\delta v_{\phi m}, \\
  &-\ii\sbar\delta v_{Rm} - 2\Omega \delta v_{\phi m} = -
  c_s^2(R)\frac{d}{dR}\left(\frac{\delta\Sigma_m}{\Sigma}\right) - \frac{d}{dR}\delta\Phi_m,\\
  & -\ii\sbar\delta v_{\phi m} + \frac{\kappa^2}{2\Omega}\delta v_{Rm} =
  -\frac{\ii m }{R}\left(c_s^2\frac{\delta\Sigma_m}{\Sigma} + \delta\Phi_m\right),
\end{align}
where $\sbar = \sigma - m\Omega$ and $\kappa^2 =
R^{-3}\p_R(R^4\Omega^2)$ is the square of the epicycle frequency. A
locally isothermal equation of state has been assumed.    
The linearised Poisson equation is 
\begin{align}
  \nabla^2\delta\Phi_m = 4\pi G \delta\Sigma_m \delta(z). 
\end{align}

The linearised equations can be combined to a single
integro-differential equation eigenvalue problem. We defer a full
exploration of the linear problem to a future study. Here, we list some
general  properties of the linear perturbations that will be useful
for analysing our simulation results.

\subsection{Global angular momentum conservation for linear
  perturbations} 
It can be shown that linear perturbations with $\phi$-dependence in the form
$\exp{(\ii m\phi)}$ satisfies angular momentum conservation
in the form 
\begin{align} 
  \frac{\p \jlin}{\p t} + \nabla\cdot\bm{F} = \tbg, 
\end{align}
\citep[e.g.][]{narayan87,ryu92,lin93} where
\begin{align}\label{lin_ang_mom_cons}
  \jlin \equiv
  -\frac{m\Sigma}{2}\imag\left(\bm{\xi}^*\cdot\frac{\p\bm{\xi}}{\p
      t} + \Omega \hat{\bm{k}}\cdot\bm{\xi}\times\bm{\xi}^* + \ii
    m \Omega |\bm{\xi}|^2\right) 
\end{align}
is the angular momentum density, $\bm{\xi}$ is the Lagragian
displacement, and $\bm{F}$ is the angular momentum flux
consisting of a Reynolds stress and a gravitational stress. Explicit
expressions for $\bm{\xi}$ can be found in, e.g. \cite{papaloizou85},
and $\bm{F}$ can be read off Eq. \ref{angflux1}.  
% \cite{papaloizou85} and \cite{lin11} respectively, and

@@

Given the real frequency $\omega$ or pattern speed $\Omega_p\equiv
\omega/m$, Eq. \ref{dispersion} can be solved 
for $|k|$, 
\begin{align}\label{wavenumber}
  |k| = \frac{\pi G \Sigma}{c_s^2}\left[1 \pm \sqrt{1 -
      Q^2(1-\nu^2)}\right], 
\end{align}
where 
% \begin{align}
%   k_T \equiv \frac{\kappa^2}{2\pi G \Sigma}
% \end{align}
% is a characteristic wavenumber, 
\begin{align}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}
\end{align}
is the usual Toomre parameter, and
\begin{align}
  \nu \equiv \frac{(\omega - m\Omega)}{\kappa}
\end{align}
is a dimensionless frequency. The upper (lower) signs in
Eq. \ref{wavenumber} correspond to short (long) waves, and $k>0$
($k<0$) correspond to trailing (leading) spirals.

>>>>

\subsubsection{Three-dimensionality}
In 3D simulations we measure the importance of vertical motion with
$\Theta$, where 
\begin{align}
  \Theta^2 \equiv \frac{\avg{v_z^2}}{\avg{v_R^2}+\avg{v_\phi^2}}, 
\end{align}
and $\avg{\cdot}$ denotes the density-weighted average, e.g., 
\begin{align}
  \avg{v_z^2} \equiv\frac{
    \int_{R_1}^{R_2}\int_{\theta_\mathrm{min}}^{\pi/2} \int_{0}^{2\pi}
    \rho v_z^2 dV}{
    \int_{R_1}^{R_2}\int_{\theta_\mathrm{min}}^{\pi/2} \int_{0}^{2\pi}
    \rho dV
  },
\end{align}
and similarly for the horizontal velocities. The radial range of
integration is taken over $r\in[R_1,R_2]$ since this is where
we find the perturbations to be confined.

@@

In a linear analysis, one assumes a steady axisymmetric background state, such as
that defined in \S\ref{model}, which is then perturbed such that
\begin{align}  
  \Sigma \to \Sigma(R) + \delta\Sigma_m(R)\exp{\left[\ii\left(-\sigma t +
        m\phi\right)\right]}, 
\end{align}
and similarly for other variables. We take $m>0$. Since the background is
axisymmetric, we identify $\delta\Sigma_m(R)\exp{\left(-\ii\sigma
    t\right)} = \Sigma_m$ as defined for diagnostics. 
The linearised mass and momentum equations are
\begin{align}
  &-\ii\sbar \delta\Sigma_m = -\frac{1}{R}\frac{d}{dR}\left(R\Sigma\delta
    v_{Rm}\right) - \frac{\ii m \Sigma}{R}\delta v_{\phi m}, \\
  &-\ii\sbar\delta v_{Rm} - 2\Omega \delta v_{\phi m} = -
  c_s^2(R)\frac{d}{dR}\left(\frac{\delta\Sigma_m}{\Sigma}\right) - \frac{d}{dR}\delta\Phi_m,\\
  & -\ii\sbar\delta v_{\phi m} + \frac{\kappa^2}{2\Omega}\delta v_{Rm} =
  -\frac{\ii m }{R}\left(c_s^2\frac{\delta\Sigma_m}{\Sigma} + \delta\Phi_m\right),
\end{align}
where $\sbar = \sigma - m\Omega$ and $\kappa^2 =
R^{-3}\p_R(R^4\Omega^2)$ is the square of the epicycle frequency. A
locally isothermal equation of state has been assumed.    
The linearised Poisson equation is 
\begin{align}
  \nabla^2\delta\Phi_m = 4\pi G \delta\Sigma_m \delta(z). 
\end{align}

The linearised equations can be combined into a single
integro-differential equation eigenvalue problem. We defer a full
exploration of the linear problem to a future study. Here, we list some
general  properties of the linear perturbations that will be useful
for describing simulation results.

\subsection{Global angular momentum conservation for linear
  perturbations} 
It can be shown that linear perturbations with $\phi$-dependence in the form
$\exp{(\ii m\phi)}$ satisfies angular momentum conservation
in the form 
\begin{align} 
  \frac{\p \jlin}{\p t} + \nabla\cdot\bm{F} = \tbg, 
\end{align}
\citep[e.g.][]{narayan87,ryu92,lin93b} where
\begin{align}\label{lin_ang_mom_cons}
  \jlin \equiv
  -\frac{m\Sigma}{2}\imag\left(\bm{\xi}^*\cdot\frac{\p\bm{\xi}}{\p
      t} + \Omega \hat{\bm{k}}\cdot\bm{\xi}\times\bm{\xi}^* + \ii
    m \Omega |\bm{\xi}|^2\right) 
\end{align}
is the angular momentum density, $\bm{\xi}$ is the Lagragian
displacement, and $\bm{F}$ is the angular momentum flux
consisting of a Reynolds stress and a gravitational stress. Explicit
expressions for $\bm{\xi}$ can be found in, e.g. \cite{papaloizou85},
and $\bm{F}$ can be read off Eq. \ref{angflux1}.  
% \cite{papaloizou85} and \cite{lin11} respectively, and

@@

Considering a neutral mode, given the real frequency $\omega$ or
pattern speed $\Omega_p\equiv 
\omega/m$, Eq. \ref{dispersion} can be solved 
for $|k|$, 
\begin{align}\label{wavenumber}
  |k| = \frac{\pi G \Sigma}{c_s^2}\left[1 \pm \sqrt{1 -
      Q^2(1-\nu^2)}\right], 
\end{align}
where 
% \begin{align}
%   k_T \equiv \frac{\kappa^2}{2\pi G \Sigma}
% \end{align}
% is a characteristic wavenumber, 
\begin{align}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}
\end{align}
is the usual Toomre parameter, and
\begin{align}
  \nu \equiv \frac{(\omega - m\Omega)}{\kappa}
\end{align}
is a dimensionless frequency. In
Eq. \ref{wavenumber}, the upper (lower) sign correspond to short
(long) waves, and $k>0$ ($k<0$) correspond to trailing (leading)
spirals.

### 2014-11-06 paper/paper.tex
appendix and minor updates
update
minor edits

\newcommand{\p}{\partial}
\newcommand{\mnras}{MNRAS}
\newcommand{\apj}{ApJ}
\newcommand{\aap}{A\&A}
\newcommand{\apjl}{ApJL}
\newcommand{\araa}{ARAA}
\newcommand{\dd}{\delta}
\newcommand{\adot}{\dot{a}}
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\gtrsim}{\;\raisebox{-.8ex}{$\buildrel{\textstyle>}\over\sim$}\;}
\newcommand{\lesssim}{\; \raisebox{-.8ex}{$\buildrel{\textstyle<}\over\sim$}\;}
\newcommand{\anrev}{{\it ARA\&A, }}
\newcommand{\apjs}{{\it ApJS, }}
\newcommand{\icar}{{\it Icarus, }}
\newcommand{\mnr}{{\it MNRAS, }}
\newcommand{\nat}{{\it Nature, }}
\newcommand{\sci}{{\it Sci, }}
\newcommand{\ana}{{\it A\&A, }}
\newcommand{\anas}{{\it A\&AS, }}
\newcommand{\aaps}{{\it A\&AS, }}
\newcommand{\anar}{{\it A\&AR, }}
\newcommand{\prd}{{\it Phys. Rev. D }}
\newcommand{\qjl}{{\it QJRAS, }}
\newcommand{\sbar}{\bar{\sigma}}
\newcommand{\avg}[1]{\langle #1 \rangle_\phi}
\newcommand{\bu}{\bm{u}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\mmax}{m_\mathrm{max}}
\newcommand{\trel}{t_\mathrm{rel}}
\newcommand{\zeus}{{\tt ZEUS-MP }}
\newcommand{\ii}{\mathrm{i}}
\newcommand{\bv}{\bm{v}}
\newcommand{\rin}{r_\mathrm{in}}
\newcommand{\rout}{r_\mathrm{out}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\tbeta}{\tilde{\beta}}
\newcommand{\teta}{\tilde{\eta}}
\newcommand{\tcool}{t_\mathrm{cool}}
\newcommand{\jlin}{j_\mathrm{lin}}
\newcommand{\jlintot}{J_\mathrm{lin}}
\newcommand{\tbg}{T_\mathrm{BG}}
\DeclareMathOperator{\erf}{erf}
\DeclareMathOperator{\real}{Re}
\DeclareMathOperator{\imag}{Im}

@@

\maketitle
\begin{abstract} 
  We present numerical simulations of astrophysical discs with an 
  imposed temperature profile and in which a sub-portion of it is
  self-gravitating. This might represent, for example, mass  
  built-up in a circumstellar or protoplanetary
  disc at the outer edge of a `dead zone', in which there
  is inefficient magneto-hydrodynamic turbulence to enable accretion. 
  We conduct numerical gravito-hydrodynamic experiments in 2D using
  the FARGO code and in 3D with the ZEUS-MP and PLUTO codes. In such
  systems we observe the development of a 
  spiral mode with azimuthal wavenumber $m=1$ with co-rotation radius 
  outside the bulk of the spiral arm, resulting in a nearly stationary 
  one-arm spiral pattern in the self-gravitating region. This one-arm 
  spiral does not require the movement of  
  the central star, as found in previous studies. Instead, we find it 
  to be associated with the imposed 
  temperature gradient and the radial disc structure. 
  We apply results from existing linear theory and attribute the
  growth of this spiral to the non-conservation of angular momentum
  for linear perturbations in discs with a forced temperature
  gradient. This enables angular momentum exchange between the spiral
  and the background disc, and may lead to instability if the  
  interaction increases the amplitude of the linear perturbations.

>>>>

\newcommand{\p}{\partial}
\newcommand{\mnras}{MNRAS}
\newcommand{\apj}{ApJ}
\newcommand{\aap}{A\&A}
\newcommand{\apjl}{ApJL}
\newcommand{\araa}{ARAA}
\newcommand{\dd}{\delta}
\newcommand{\adot}{\dot{a}}
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\gtrsim}{\;\raisebox{-.8ex}{$\buildrel{\textstyle>}\over\sim$}\;}
\newcommand{\lesssim}{\; \raisebox{-.8ex}{$\buildrel{\textstyle<}\over\sim$}\;}
\newcommand{\anrev}{{\it ARA\&A, }}
\newcommand{\apjs}{{\it ApJS, }}
\newcommand{\icar}{{\it Icarus, }}
\newcommand{\mnr}{{\it MNRAS, }}
\newcommand{\nat}{{\it Nature, }}
\newcommand{\sci}{{\it Sci, }}
\newcommand{\ana}{{\it A\&A, }}
\newcommand{\anas}{{\it A\&AS, }}
\newcommand{\aaps}{{\it A\&AS, }}
\newcommand{\anar}{{\it A\&AR, }}
\newcommand{\prd}{{\it Phys. Rev. D }}
\newcommand{\qjl}{{\it QJRAS, }}
\newcommand{\sbar}{\bar{\sigma}}
\newcommand{\avg}[1]{\langle #1 \rangle}
\newcommand{\bu}{\bm{u}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\mmax}{m_\mathrm{max}}
\newcommand{\trel}{t_\mathrm{rel}}
\newcommand{\zeus}{{\tt ZEUS-MP }}
\newcommand{\ii}{\mathrm{i}}
\newcommand{\bv}{\bm{v}}
\newcommand{\rin}{r_\mathrm{in}}
\newcommand{\rout}{r_\mathrm{out}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\tbeta}{\tilde{\beta}}
\newcommand{\teta}{\tilde{\eta}}
\newcommand{\tcool}{t_\mathrm{cool}}
\newcommand{\jlin}{j_\mathrm{lin}}
\newcommand{\jlintot}{J_\mathrm{lin}}
\newcommand{\tbg}{T_\mathrm{BG}}
\DeclareMathOperator{\erf}{erf}
\DeclareMathOperator{\real}{Re}
\DeclareMathOperator{\imag}{Im}

@@

\maketitle
\begin{abstract} 
  We present numerical simulations of astrophysical discs with an 
  imposed temperature profile and in which a sub-portion of it is
  self-gravitating. This might represent, for example, mass  
  built-up in a circumstellar or protoplanetary
  disc at the outer edge of a `dead zone', in which there
  is inefficient magneto-hydrodynamic turbulence to enable accretion. 
  We conduct numerical gravito-hydrodynamic experiments in 2D using
  the FARGO code and in 3D with the ZEUS-MP and PLUTO codes. In such
  discs we observe the development of a 
  spiral mode with azimuthal wavenumber $m=1$ with co-rotation radius 
  outside the bulk of the spiral arm, resulting in a nearly stationary 
  one-arm spiral pattern in the self-gravitating region of the
  disc. The development of this one-arm 
  spiral does not require the movement of  
  the central star, as found in previous studies. Instead, we find it 
  to be associated with the imposed 
  temperature gradient and the radial disc structure. 
  We apply results from existing linear theory and attribute the
  growth of this spiral to the non-conservation of angular momentum
  for non-axisymmetric linear perturbations in discs with a forced temperature
  gradient. Such interaction may lead to an unstable exchange of angular momentum between the spiral 
  and the background disc. However, because the exchange mechanism does
  not explicitly depend on self-gravity, we speculate it may also
  enhance low-frequency one-arm spirals in non-self-gravitating
  discs.

### 2014-11-06 paper/results.tex
minor edits

\subsection{Trapping the low-frequency $m=1$ spiral}\label{fargo_m1}
In order to focus on the $m=1$ spiral that eventually emerges, here 
we examine a simulation initialised with only $m=1$
perturbations, by setting $M=1$ in Eq. \ref{randpert}. The disc
evolution is largely similar to the previous case, but the final spiral is
more coherent, which facilitates the analysis below.

Fig. \ref{2d_fargo_viz} shows the $m=1$ surface density
at the end of this simulation. For this one-arm spiral
we measure a co-rotation radius and growth rate of 
\begin{align*}
  &R_c \simeq 4.4R_0,\\
  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_k(R_c). 
\end{align*}
The spiral can be considered as low frequency because its 
pattern speed $\Omega_p\lesssim 0.3\Omega$ in $R\in[R_{1},R_{2}]$,
where it is  most prominent. The growth rate $\gamma$ is also slow
relative to the local rotation. Thus, the spiral 
pattern is nearly stationary.

@@

Next, we write $\Sigma_1 = 
|\Sigma_1|\exp{(\ii kR)}$, where $k$ is real, and assume the amplitude
$|\Sigma_1|$ varies slowly compared to the complex phase. This is the
main assumption in local theory. We plot in
Fig. \ref{fargo_wavenumber} the normalised radial wavenumber associated with
the $m=1$ spiral. We find 
\begin{align*}
  kR \sim \frac{\pi G \Sigma}{c_s^2}R \sim \frac{1}{hQ}, 
\end{align*}
where we used $Q\sim c_s\Omega/\pi G \Sigma$ and $R\Omega/c_s\sim
h^{-1}$. Since $Q=O(1)$ and $h\ll 1$ imply $|kR|\gg 1$, we 
can apply results from local theory self-consistently. Recall $k>0$
means the spiral is trailing, as seen in Fig. \ref{2d_fargo_viz}.

@@

Using the estimated value of $R_c$, we plot in
Fig. \ref{fargo_qbarrier} the quantity $\nu^2 - 1 + Q^{-2}$, which is
required to be positive in local theory for purely wave-like
solutions to the dispersion relation (Eq. \ref{dispersion}) when the
mode frequency is given. The figure shows two $Q$-barriers located in
the inner disc, at $R_{Qb}=R_0$ and $R_{Qb}=1.6R_0$, which is indeed the
region where the $m=1$ spiral develops. This suggests that the one-arm
spiral is trapped. There is one outer Lindblad resonance at $R_L\simeq
7.2R_0$. Thus, waves may be launched in $R\gtrsim 7.2R_0$ by the main 
spiral disturbance in the inner disc \citep{lin11b}. 
%this would be a de-stabilising
% effect.

@@

In order to track down the origin for the (slow) growth of the
$m=1$ spiral, we invoke angular momentum conservation for linear
perturbations, Eq. \ref{lin_ang_mom_cons}. Assuming fluxes are
negligible at the disc boundaries, we can integrate
Eq. \ref{lin_ang_mom_cons} to give
\begin{align}\label{baroclinic_torque_int}
  \frac{d}{dt}\underbrace{\int_{R_\mathrm{min}}^{R_\mathrm{max}}\jlin
    2\pi R dR}_{\jlintot} 
  =\int_{R\mathrm{min}}^{R\mathrm{max}}T_\mathrm{BG} 2\pi R dR, 
\end{align}
where we recall $T_{BG}$ is the torque density related to the imposed
sound-speed profile (Eq. \ref{baroclinic_torque}). We have explicitly
computed both sides of Eq. \ref{baroclinic_torque_int} using
simulation data, and compare them in Fig. \ref{fargo_angmom_ex}. There
is a good match between the two torques, especially at early times
$t\lesssim110P_0$. The match is less good later on, when the spiral
amplitude is no longer small ($\Delta\Sigma_1\sim 0.2$ at $t=110P_0$
and $\Delta\Sigma_1\sim 0.4$ by $t=120P_0$) and linear theory becomes
less valid. Fig. \ref{fargo_angmom_ex} confirms that angular momentum
is lost from the wave to the disc.

@@

Fig. \ref{fargo_varq} compares the $m=1$ spiral amplitudes as a
function of $q$. We indeed observe slower growth with decreasing
$q$. Although the figure indicates growth for the strictly isothermal
disc ($q=0$), we did not observe a coherent one-arm spiral upon
inspection of the $m=1$ surface density field. The growth in this case
may be associated with higher $m$ modes. %"non-linear effects"

@@

We plot growth rates of the $m=1$ spiral as a function of $q$ in 
Fig. \ref{fargo_varq_growth}. The correlation is well-fitted by a
linear relation
\begin{align*}
  \gamma \simeq \left[0.015 q - 7.9\times10^{-4}\right] \Omega_k(R_0). 
\end{align*}
This confirms that the temperature gradient is responsible for the
development of the one-arm spiral observed in our simulations. 
\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ratemax_fargo_varq.ps}    
  \caption{Growth rates of the $m=1$ spiral mode as a function of the
    imposed sound-speed gradient $q$ (asterisks). A linear fit is also
    plotted (dotted line). 
    \label{fargo_varq_growth}} 
\end{figure}

% only cases where a coherent spiral is seen is plotted

We also performed a series of simulations with aspect-ratio
$h\in[0.03,0.07]$ with fixed $q=1$. This also affects the temperature
gradient since $c_s \propto h$. However, with other parameters equal
to that in the fiducial simulation,  varying $h$ also changes the disc
mass. For $h\in[0.03,0.07]$ the total disc mass ranges from
$M_d=0.052M_*$ to $M_d=0.12M_*$ and the 
mass within $R\in[R_1,R_2]$  ranges from $0.033M_*$ to $0.062M_*$.

Fig. \ref{fargo_varh_growth} shows the growth rates of the $m=1$
spiral in $R\in[R_1,R_2]$ as a function of $h$. Growth rates increases
with $h$, roughly as 
\begin{align}
  \gamma \simeq \left[0.10h + 8.3\times10^{-3}\right]\Omega_k(R_0).  
\end{align}
However, the linear fit is less good than for variable $q$ above. This
may be due to the change in the total disc mass when $h$ changes. We
find no qualitative difference between the spiral pattern that
emerges.

>>>>

\subsection{Trapping the low-frequency $m=1$ spiral}\label{fargo_m1}
In order to focus on the $m=1$ spiral that eventually emerges, from
now on we examine simulations initialised with
$M=1$ in Eq. \ref{randpert}. The disc 
evolution is largely similar to the previous case, but the final spiral is
more coherent, which facilitates the analysis below.

Fig. \ref{2d_fargo_viz} shows the $m=1$ surface density
at the end of this simulation. For this one-arm spiral
we measure a co-rotation radius and growth rate of 
\begin{align*}
  &R_c \simeq 4.4R_0,\\
  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_k(R_c). 
\end{align*}
The spiral can be considered as low frequency because its 
pattern speed $\Omega_p\lesssim 0.3\Omega$ in $R\in[R_{1},R_{2}]$,
where it is  most prominent. The growth rate $\gamma$ is also slow
relative to the local rotation. (Although the characteristic growth
time $\gamma^{-1} \simeq 10P_0$.) Thus, the spiral 
pattern is nearly stationary.

@@

Next, we write $\Sigma_1 = 
|\Sigma_1|\exp{(\ii kR)}$, where $k$ is real, and assume the amplitude
$|\Sigma_1|$ varies slowly compared to the complex phase. This is the
main assumption in local theory. We measure $k$ numerically and plot
its normalised value in Fig. \ref{fargo_wavenumber}. We find 
\begin{align*}
  kR \sim \frac{\pi G \Sigma}{c_s^2}R \sim \frac{1}{hQ}, 
\end{align*}
where we used $Q\sim c_s\Omega/\pi G \Sigma$ and $R\Omega/c_s\sim
h^{-1}$. Since $Q=O(1)$ and $h\ll 1$ imply $|kR|\gg 1$, we 
can apply results from local theory self-consistently. Recall $k>0$
corresponds to a trailing spiral, as seen in Fig. \ref{2d_fargo_viz}.

@@

Using the estimated value of $R_c$, we plot in
Fig. \ref{fargo_qbarrier} the quantity $\nu^2 - 1 + Q^{-2}$, which is
required to be positive in local theory for purely wave-like
solutions to the dispersion relation (Eq. \ref{dispersion}) when the
mode frequency is given. The figure shows two $Q$-barriers located in
the inner disc, at $R_{Qb}=R_0$ and $R_{Qb}=1.6R_0$, which is indeed the
region where the $m=1$ spiral develops. This suggests that the one-arm
spiral is trapped. There is one outer Lindblad resonance at $R_L\simeq
7.2R_0$. Thus, waves may be launched in $R\gtrsim 7.2R_0$ by the  
spiral disturbance in the inner disc \citep{lin11b}. 
%this would be a de-stabilising
% effect.  we discuss this in discussions?

@@

In order to track down the origin for the (slow) growth of the
$m=1$ spiral, we invoke angular momentum conservation for linear
perturbations, Eq. \ref{lin_ang_mom_cons}. Assuming fluxes are
negligible at the disc boundaries, we can integrate
Eq. \ref{lin_ang_mom_cons} to give
\begin{align}\label{baroclinic_torque_int}
  \frac{d}{dt}\underbrace{\int_{R_\mathrm{min}}^{R_\mathrm{max}}\jlin
    2\pi R dR}_{\jlintot} 
  =\int_{R\mathrm{min}}^{R\mathrm{max}}T_\mathrm{BG} 2\pi R dR, 
\end{align}
where we recall $T_\mathrm{BG}$ is the torque density due to the imposed
sound-speed profile (Eq. \ref{baroclinic_torque}). We have explicitly
computed both sides of Eq. \ref{baroclinic_torque_int} using
simulation data, and compare them in Fig. \ref{fargo_angmom_ex}. There
is a good match between the two torques, especially at early times
$t\lesssim110P_0$. The match is less good later on, when the spiral
amplitude is no longer small ($\Delta\Sigma_1\sim 0.2$ at $t=110P_0$
and $\Delta\Sigma_1\sim 0.4$ by $t=120P_0$) and linear theory becomes
less valid. Fig. \ref{fargo_angmom_ex} confirms that the $m=1$ spiral
wave experiences a negative torque that further reduces its angular
momentum. This is consistent with angular momentum component
measurements (Fig. \ref {2d_angmom}).

@@

Fig. \ref{fargo_varq} compares the $m=1$ spiral amplitudes as a
function of $q$. We indeed observe slower growth with decreasing
$q$. Although the figure indicates growth for the strictly isothermal
disc ($q=0$), we did not observe a coherent one-arm spiral upon
inspection of the $m=1$ surface density field. The growth in this case
may be associated with higher $m$ modes.  
% show result of isothermal calc?

@@

We plot growth rates of the $m=1$ spiral as a function of $q$ in 
Fig. \ref{fargo_varq_growth}. The correlation can be fitted with a
linear relation
\begin{align*}
  \gamma \simeq \left[0.015 q - 7.9\times10^{-4}\right] \Omega_k(R_0). 
\end{align*}
This confirms that the temperature gradient is responsible for the
development of the one-arm spiral observed in our simulations. 
\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ratemax_fargo_varq.ps}    
  \caption{Growth rates of the $m=1$ spiral mode as a function of the
    imposed sound-speed gradient $q$ (asterisks). A linear fit is also
    plotted (dotted line). 
    \label{fargo_varq_growth}} 
\end{figure}
%maybe extend the low q simulations to get better rates?

We also performed a series of simulations with variable aspect-ratio
$h\in[0.03,0.07]$ but fixed $q=1$. This also affects the temperature
gradient since $c_s \propto h$. However, with other parameters equal
to that in the fiducial simulation,  varying $h$ also changes the disc
mass. For $h\in[0.03,0.07]$ the total disc mass ranges from
$M_d=0.052M_*$ to $M_d=0.12M_*$ and the 
mass within $R\in[R_1,R_2]$  ranges from $0.033M_*$ to $0.062M_*$. 
% similar Q profiles

Fig. \ref{fargo_varh_growth} shows the growth rates of the $m=1$
spiral in $R\in[R_1,R_2]$ as a function of $h$. Growth rates increases
with $h$, roughly as  
\begin{align}
  \gamma \simeq \left[0.10h + 8.3\times10^{-3}\right]\Omega_k(R_0).  
\end{align}
However, a linear fit is less good than for variable $q$ above. This
may be due to the change in the total disc mass when $h$ changes. We
find no qualitative difference between the spiral pattern that
emerges.
